---
title: "Central Tendency Variability & Descriptive Statistics"
author: "Ben Gonzalez"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 5
    collapsed: false
    smooth_scroll: true
---

```{r setup, klippy, echo=FALSE, include=TRUE,message=FALSE}
klippy::klippy(tooltip_message = "Copy Code",tooltip_success = "Copy Successful!!!",position = "right")
library(mathjaxr)
library(latexpdf)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}

gonzo_theme <- function() {
  list(
    labs(
      caption = "Mtcars Data Set: 1974 Motor Trend US"),
    theme(
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.caption = element_text(color = "gray12",size = 10)
    )
  )
}


```

### Descriptive Statistics

### Measures of Central Tendency

#### Mean

The mean or average and sometimes referred to as the arithmeic mean (Hatcher 2013)

We take the sum $\sum$ of all the scores $X$ in a column and then divide them by the total number of scores or $N$

Assumptions to calculate the mean:

+ The variable is quantitative and is captured via an interval or ratio scale.
+ The distribution of the data is approx. normal or symmetrical

The formula for the mean is as follows:

$$
\overline{X}=\frac{\sum{X}}{N}
$$
#### Median

According to Hatcher(2013) the median is the score at which 50% of the distribution falls at or below this score.

$Odd:$

One way to calculate the median is to order the values in the data and order them from largest to smallest. The middle most number is the $median$.

$Even$

If you have an even number of data. You order the data again from largest to smallest and then you take the two middle numbers in your data and divide them by $2$.

You can also utilize the ``median()` function in `R`

```{r,warning=FALSE,message=FALSE}
median(mtcars$mpg)
```

Assumptions to calculate the median:

+ The variable is assessed on an ordinal scale
+ The variable is assess on an interval or ratio scale

#### Mode

The mode is the most frequently occurring number in a distribution.

There are two types of modes `unimodal` and `bimodal`. 

$unimodal$ - there is only one mode present in the distribution.

$bimodal$ - there are two modes present in a distribution.

Assumptions to calculate the mode:

+ Typically used with variables assessed on a nominal scale; e.g. categorical data


```{r ,warning=FALSE,message=FALSE}
# Create the function.
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

getmode(mtcars$cyl)

# hist(mtcars$cyl)


# Create histogram for 'cyl' column
hist_obj <- hist(mtcars$cyl, 
                 main = "Histogram of Number of Cylinders (cyl)", 
                 xlab = "Number of Cylinders", 
                 ylab = "Frequency", 
                 col = "lightblue", 
                 border = "black", 
                 plot = FALSE)  # Don't plot the histogram yet, just store the data

# Plot the histogram
plot(hist_obj, 
     main = "Histogram of Number of Cylinders (cyl)", 
     xlab = "Number of Cylinders", 
     ylab = "Frequency", 
     col = "lightblue", 
     border = "black")

# Add counts on top of each bar
text(x = hist_obj$mids, 
     y = hist_obj$counts, 
     labels = hist_obj$counts, 
     pos = 1.5,      # Position text above the bars
     col = "black", # Color of the text
     cex = 0.8)     # Size of the text


# ggplot(mtcars, aes(x = cyl)) +
#   geom_histogram(fill = "lightblue", color = "blue") +
#   labs(title = "Variation of MPG in Mtcars Data Set", x = "MPG Variance") +
#   theme_minimal()+gonzo_theme()

```


### Measures of Variability


### Variance 

$Measure\;of\;variability$ - indicates the amount of dispersion displayed by a set of scores. (Hatcher 2013)

$Variance$ - measure of variability computed by taking the average of the squared deviations from the mean. (Hatcher 2013) 

Variance can be calculated via the following formula. We first calculate the average distance from each data point from the mean of our data.

<!-- https://campus.datacamp.com/courses/introduction-to-statistics-in-r/summary-statistics?ex=7 -->


The formula for calculating the $Sample$ variance is as follows:

$$
S^2\;=\;\frac{\sum(X_{i}\;-\;{\overline{X}})^2}{N\;-1}
$$

```{r mtcars-variance-manual-calculation, echo=FALSE}

###Number of observations in our data set
N <- 32

variance_sum_one<- sum((mtcars$mpg-mean(mtcars$mpg))^2)/(N-1)

```

The variance of MPG in the Mtcars data set is `r variance_sum_one`

***

You can also utilize the `var()` function in R to calculate the variance. 

```{r}
var(mtcars$mpg)
```


***

#### Visualization of variation in a data set

One way to look at the variance without needing to calculate it is by utilizing a boxplot or a histogram. 


##### Boxplot

```{r,warning=FALSE,message=FALSE,fig.align='center'}
# Boxplot of mpg in mtcars
ggplot(mtcars, aes(x = "", y = mpg)) +
  geom_boxplot(fill = "lightblue", color = "blue") +
  labs(title = "Variation of MPG in mtcars", y = "Miles Per Gallon (MPG)", x = "MPG Variation") +
  theme_minimal()+gonzo_theme()

```


The boxplot for **MPG** in the mtcars data set visualizes the variation in fuel efficiency across different cars. Here's what we can interpret from the boxplot:

+ **Median (Center Line):** The thick horizontal line in the middle of the box represents the median value of MPG, which is the middle value when all the MPG values are ordered from smallest to largest. This gives us a sense of the "central tendency" or typical value for MPG in this dataset.

+ **Interquartile Range (IQR):** The box represents the interquartile range (IQR), which includes the middle 50% of the data. The top and bottom edges of the box represent the third quartile (Q3) and first quartile (Q1), respectively. This range gives us an idea of the spread of the central half of the data.

+ **Whiskers:** The lines extending from the box (called whiskers) show the range of values within 1.5 times the IQR from the quartiles. These lines help to visualize the spread of data, excluding outliers.

+ **Outliers:** Points that fall outside of the whiskers (typically shown as individual dots) are considered outliers. These are values that are significantly different from the rest of the data. For example, the MPG values for cars with extreme fuel efficiency or inefficiency could appear as outliers in this plot.



##### Histogram

```{r variance-histogram,warning=FALSE,message=FALSE}
# Histogram to visualize the distribution (and variance) of mpg
hist(mtcars$mpg, 
     main = "Histogram of mpg (Variance)", 
     xlab = "Miles per Gallon", 
     col = "lightgreen", 
     breaks = 10)

```


##### Density Plot

```{r variance-density-plot,warning=FALSE,message=FALSE}
# Density plot to visualize the spread (variance) of mpg
# plot(density(mtcars$mpg), 
#      main = "Density Plot of mpg (Variance)", 
#      xlab = "Miles per Gallon", 
#      col = "darkblue", 
#      lwd = 2)


ggplot(mtcars, aes(x = mpg)) +
  geom_density(fill = "lightblue", color = "blue") +
  stat_function(fun = dnorm, args = list(mean = mean(mtcars$mpg), sd = sd(mtcars$mpg)), 
                color = "red", linetype = "dashed", size = 1.5) + 
  labs(title = "Variation of MPG in Mtcars Data Set", x = "MPG Variance") +
  theme_minimal()+gonzo_theme()



```

#### Quartiles


Quartiles is when the data is split up into `4` equal parts `25%` each. 

The syymbols used for `quartiles` are as follows.

***

$$Q1\;=\;First\;Quartile\\
Q2\;=Second\;Quartile\\
Q3\;=Third\;Quartile\\
Q4\;=Fourth\;Quartile $$

***

We can easily compute the `quartiles` in `R` using the `quantile` function.

```{r quartiles,warning=FALSE,message=FALSE}
quantile(mtcars$mpg)
```
We can thus compute the interquartle range or $IQR$ as follows.

$IQR\;=\;Q_{3}-Q_{1}$

In our data above this would be the $Q_{3}$`75%` - $Q_{1}$`25%` or `r 22.800-15.425`

```{r iqr,warning=FALSE,message=FALSE}
IQR(mtcars$mpg)
```

