---
title: "SPSS Syntax"
output:
  html_document:
    code_folding: show
---

```{r setup, klippy, echo=FALSE, include=TRUE}
klippy::klippy(tooltip_message = "Copy Code",tooltip_success = "Copy Successful!!!",position = "right")
library(mathjaxr)
library(latexpdf)
knitr::opts_chunk$set(echo = TRUE)
```


```{r,eval=FALSE}
GET DATA
  /TYPE=XLSX
  /FILE='C:\path\to\yourfile.xlsx'
  /SHEET=name 'Sheet1'
  /CELLRANGE=FULL
  /READNAMES=ON.
```



```{r,eval=FALSE}
GET DATA
  /TYPE=TXT
  /FILE='C:\path\to\yourfile.csv'
  /DELCASE=LINE
  /DELIMITERS=","
  /ARRANGEMENT=DELIMITED
  /FIRSTCASE=2
  /VARIABLES=var1 F8.2 var2 F8.2 var3 F8.2.
```

### T-test code

```{r,eval=FALSE}
T-TEST GROUPS=gender(1 2)
  /VARIABLES=math_score
  /CRITERIA=CI(.95).
```

### Z-score code

```{r,eval=FALSE}

DESCRIPTIVES VARIABLES=ALL /SAVE

```

### Chi-square code

Below is the code for calculating the chi-square statistic.

```{r,eval=FALSE}
CROSSTABS 
  /TABLES=Mjob BY internet  
  /STATISTICS=CHISQ
  /CELLS=COUNT EXPECTED.
```

### Pearson Correlation

Below is the code for calculating the Pearson correlation.

```{r,eval=FALSE}
CORRELATIONS 
  /VARIABLES=var1 var2
  /STATISTICS=DESCRIPTIVES.
```


### Spearman Rank Correlation

```{r,eval=FALSE}
NONPAR CORR 
  /VARIABLES=studytime traveltime
  /PRINT=SPEARMAN TWOTAIL NOSIG FULL
  /MISSING=PAIRWISE.
```

### Phi-coefficient

Below is the syntax for calculating the phi-coefficient statistic.

```{r,eval=FALSE}
CROSSTABS 
  /TABLES=sex BY internet 
  /FORMAT=AVALUE TABLES 
  /STATISTICS=PHI 
  /CELLS=COUNT 
  /COUNT ROUND CELL.

```

<!-- ![](./Graphs/phi-coefficient-sex-by-internet-student-data-csv.pdf) -->

### Partial Correlation

Below is the code for calculating the partial correlation.

```{r,eval=FALSE}
PARTIAL CORR 
  /VARIABLES=hp mpg BY cyl
  /SIGNIFICANCE=TWOTAIL
  /MISSING=LISTWISE.
```

### Scatterplot in SPSS

Below is the code for generating a scatterplot in SPSS

```{r,eval=FALSE}
GRAPH 
  /SCATTERPLOT(BIVAR)=independent_variable WITH dependent_variable.
```



### Bivariate Regression

Below is the code for performing a bivariate regression in SPSS.

```{r,eval=FALSE}
REGRESSION 
  /DEPENDENT=dependent_variable
  /METHOD=ENTER independent_variable.
```
