---
title: "Z-scores"
author: "Ben Gonzalez"
date: '2024-02-25'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    collapsed: true
    smooth_scroll: true
    fig_width: 14
    fig_height: 5
---

```{r setup, klippy, echo=FALSE,message=FALSE}
klippy::klippy(tooltip_message = "Copy Code",tooltip_success = "Copy Successful!!!",position = "right")
library(dplyr)
library(ggplot2)
library(reactable)
library(gridExtra)

knitr::opts_chunk$set(echo = TRUE)
```


```{r,echo=FALSE,warning=FALSE,message=FALSE}

gonzo_theme <- function() {
  list(
    labs(
      caption = "Mtcars Data Set: 1974 Motor Trend US"),
    theme(
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.caption = element_text(color = "gray12",size = 10)
    )
  )
}


```

We will be utilizing the mtcars data set. Here we will normalize the mpg scores and return their equivalent z-score values.

```{r mtcars}
###Compute the last_evaluation Z scores
mtcars_all_zscores<- mtcars %>% 
  mutate(mpg_zscore = (mtcars$mpg - mean(mtcars$mpg))/sd(mtcars$mpg))

reactable(mtcars_all_zscores,filterable = T,striped = T,searchable = T,minRows = 5,defaultPageSize = 5)
```


Here we are calculating the z-scores relevant to the Camaro Z28 and the Toyota Corona. Due to the difference in everything from engine size, displacement, and mpg we can show two very contrasting vehicles.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
z_one <- as.numeric(mtcars_all_zscores[24,][12])
one_tailed_value <- pnorm(z_one ,lower.tail = TRUE)
# one_tailed_value

two_tailed_value <- 2*(1-pnorm(abs(z_one)))
# two_tailed_value

perc_distribution <- 2*pnorm(abs(z_one))-1

z_employee_one <- cbind(mtcars$mpg[24],mtcars_all_zscores[24,][12],one_tailed_value,two_tailed_value,perc_distribution)

colnames(z_employee_one) <- c("Raw Score","Z Score","One Tailed +/- Z","Two tailed +/-Z","Area between +/- Z")


z_two <- as.numeric(mtcars_all_zscores[21,][12])
one_tailed_value_two <- pnorm(z_two,lower.tail = FALSE)
# one_tailed_value_two


two_tailed_value_two <- 2*(1-pnorm(abs(z_two)))
# two_tailed_value_two

perc_distribution_two <- 2*pnorm(z_two)-1
# perc_distribution_two

z_employee_two <- cbind(mtcars$mpg[21],mtcars_all_zscores[21,][12],one_tailed_value_two,two_tailed_value_two,perc_distribution_two)
colnames(z_employee_two) <- c("Raw Score","Z Score","One Tailed +/- Z","Two tailed +/-Z","Area between +/- Z")


z_employee_scores<- rbind(z_employee_one,z_employee_two)

colnames(z_employee_scores) <- c("Raw Score","Z Score","One Tailed +/- Z","Two tailed +/-Z","Area between +/- Z")

rownames(z_employee_scores) <- c("Camaro Z28","Toyota Corona")

reactable(z_employee_scores)

```

(1) the percentage of the distribution that falls above (or below if your score is negative) that value (this is the one-tailed for -z or +z), (2) the percentage of the distribution that falls above and below the |z| (this is the two-tailed value), and (3) the percentage of the distribution that falls between +/- z (this is the area between +/- z).

### One tailed value for Z

The percentage of the distribution that falls above (or below if your score is negative) that value (this is the one-tailed for -z or +z)

***

```{r,warning=FALSE,message=FALSE}
# Set the negative one-tailed critical value (Z-value)
z <- as.numeric(mtcars_all_zscores[20,][12])  # Example value, you can change it as needed

# Calculate the percentage below the negative one-tailed critical value
percentage_below <- pnorm(z) * 100

percentage_below

library(ggplot2)

# Set the negative one-tailed critical value (Z-value)
z <- as.numeric(mtcars_all_zscores[24,][12])  # Example value, you can change it as needed

# Calculate the percentage below the negative one-tailed critical value
percentage_below <- pnorm(z) * 100

# Generate a sequence of values along the x-axis
x <- seq(-3, 3, length.out = 1000)

# Calculate the corresponding probabilities under the standard normal curve
y <- dnorm(x)

# Create a data frame with x and y values
df <- data.frame(x, y)

# Create the plot
ggplotone<- ggplot(df, aes(x, y)) +
  geom_line() +
  geom_ribbon(data = subset(df, x <= z), aes(ymin = 0, ymax = y), fill = "blue", alpha = 0.3) +
  geom_vline(xintercept = z, linetype = "dashed", color = "red") +
  annotate("text", x = z - 0.5, y = 0.1, label = paste0("Z =", z), color = "red") +
  theme_minimal() +
  labs(x = "Z-value", y = "Density", title = "Camaro Z28: Percentage of Distribution Below Negative One-Tailed Z-value") +
  annotate("text", x = z - 0.5, y = 0.3, label = paste0(round(percentage_below, 2), "% below"), color = "blue")+gonzo_theme()


```


```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Set the positive one-tailed critical value (Z-value)
z <- as.numeric(mtcars_all_zscores[21,][12])  # Example value, you can change it as needed

# Calculate the percentage above the positive one-tailed critical value
percentage_above <- (1 - pnorm(z)) * 100

percentage_above

library(ggplot2)

# Set the positive one-tailed critical value (Z-value)
z <- as.numeric(mtcars_all_zscores[21,][12])  # Example value, you can change it as needed

# Calculate the percentage above the negative one-tailed critical value
percentage_below <- (1 - pnorm(z)) * 100

# Generate a sequence of values along the x-axis
x <- seq(-3, 3, length.out = 1000)

# Calculate the corresponding probabilities under the standard normal curve
y <- dnorm(x)

# Create a data frame with x and y values
df <- data.frame(x, y)

# Create the plot
ggplotwo<- ggplot(df, aes(x, y)) +
  geom_line() +
  geom_ribbon(data = subset(df, x >= z), aes(ymin = 0, ymax = y), fill = "blue", alpha = 0.3) +
  geom_vline(xintercept = z, linetype = "dashed", color = "red") +
  annotate("text", x = z - 0.5, y = 0.1, label = paste0("Z =", z), color = "red") +
  theme_minimal() +
  labs(x = "Z-value", y = "Density", title = "Toyota Corona: Percentage of Distribution Above Positive One-Tailed Z-value") +
  annotate("text", x = z - 0.5, y = 0.3, label = paste0(round(percentage_below, 2), "% above"), color = "blue")+gonzo_theme()

library(gridExtra)

gridExtra::grid.arrange(ggplotone,ggplotwo,ncol=2)


```

***

### Two Tailed :Percentage above and below Z or |Z|

The percentage of the distribution that falls above and below the |z| (this is the two-tailed value)

***

```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(ggplot2)

# Define the Z-value
z <- as.numeric(mtcars_all_zscores[24,][12])  # Example value, you can change it as needed
z_abs <- abs(z)  # Absolute value of Z

# Calculate the probability in one tail
prob_one_tail <- pnorm(z_abs)

# Calculate the two-tailed value
prob_two_tailed <- 2 * (1 - prob_one_tail) * 100

# Generate a sequence of values along the x-axis
x <- seq(-3, 3, length.out = 1000)

# Calculate the corresponding probabilities under the standard normal curve
y <- dnorm(x)

# Create a data frame with x and y values
df <- data.frame(x, y)

# Create the plot
ggplotthree <- ggplot(df, aes(x, y)) +
  geom_line() +
  geom_ribbon(data = subset(df, x <= -z_abs), aes(ymin = 0, ymax = y), fill = "red", alpha = 0.3) +
  geom_ribbon(data = subset(df, x >= z_abs), aes(ymin = 0, ymax = y), fill = "blue", alpha = 0.3) +
  geom_vline(xintercept = -z_abs, linetype = "dashed", color = "black") +
  geom_vline(xintercept = z_abs, linetype = "dashed", color = "black") +
  theme_minimal() +
  labs(x = "Z-value", y = "Density", title = paste("Camaro Z28: Percentage of |Z| Distribution =", round(z_abs, 3))) +
  annotate("text", x = -z_abs - 0.5, y = 0.3, label = paste0(round((1 - prob_one_tail) * 100, 2), "% in left tail"), color = "red") +
  annotate("text", x = z_abs + 0.5, y = 0.2, label = paste0(round((1 - prob_one_tail) * 100, 2), "% in right tail"), color = "blue") +
  annotate("text", x = 0, y = 0.35, label = paste0("Total: ", round(prob_two_tailed, 2), "%"), color = "black", fontface = "bold")+gonzo_theme()



```


```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(ggplot2)

# Define the Z-value
z <- as.numeric(mtcars_all_zscores[21,][12])  # Example value, you can change it as needed
z_abs <- abs(z)  # Absolute value of Z

# Calculate the probability in one tail
prob_one_tail <- pnorm(z_abs)

# Calculate the two-tailed value
prob_two_tailed <- 2 * (1 - prob_one_tail) * 100

# Calculate the percentage below the Z-value
percentage_below <- pnorm(z_abs) * 100

# Calculate the percentage above the Z-value
percentage_above <- (1 - pnorm(z_abs)) * 100

# Generate a sequence of values along the x-axis
x <- seq(-3, 3, length.out = 1000)

# Calculate the corresponding probabilities under the standard normal curve
y <- dnorm(x)

# Create a data frame with x and y values
df <- data.frame(x, y)

# Create the plot
ggplotfour<- ggplot(df, aes(x, y)) +
  geom_line() +
  geom_ribbon(data = subset(df, x <= -z), aes(ymin = 0, ymax = y), fill = "blue", alpha = 0.3) +
  geom_ribbon(data = subset(df, x >= z), aes(ymin = 0, ymax = y), fill = "red", alpha = 0.3) +
  geom_vline(xintercept = z, linetype = "dashed", color = "black") +
  theme_minimal() +
  labs(x = "Z-value", y = "Density", title = paste("Toyota Corona: Percentage of |Z| Distribution =", round(z_abs, 3))) +
  annotate("text", x = -z - 0.5, y = 0.3, label = paste0(round(percentage_above, 2), "% below"), color = "blue") +
  annotate("text", x = z + 0.5, y = 0.2, label = paste0(round(percentage_above, 2), "% above"), color = "red") +
  annotate("text", x = 0, y = 0.35, label = paste0("Total: ", round(prob_two_tailed, 2), "%"), color = "black", fontface = "bold")+gonzo_theme()

library(gridExtra)

gridExtra::grid.arrange(ggplotthree,ggplotfour,ncol=2)

```

***

### Percentage between +/- Z

The percentage of the distribution that falls between +/- z (this is the area between +/- z)

***

```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(ggplot2)

# Set the Z-value
z <- abs(as.numeric(mtcars_all_zscores[24,][12]))  # Example Z-value

# Calculate the cumulative probabilities at the Z-values
cum_prob_positive <- pnorm(z)
cum_prob_negative <- pnorm(-z)

# Calculate the percentage between the Z-values (two-tailed)
percentage_between <- (cum_prob_positive - cum_prob_negative) * 100

# Generate a sequence of values along the x-axis
x <- seq(-3, 3, length.out = 1000)

# Calculate the corresponding probabilities under the standard normal curve
y <- dnorm(x)

# Create a data frame with x and y values for plotting
df <- data.frame(x, y)

# Create the plot with zoomed-in x-axis
ggplotfive<- ggplot(df, aes(x, y)) +
  geom_line() +
  geom_ribbon(data = subset(df, x >= -z & x <= z), aes(ymin = 0, ymax = y), fill = "blue", alpha = 0.3) +
  theme_minimal() +
  labs(x = "Z-value", y = "Density", title = "Percentage of Distribution Between +/- Z") +
  annotate("text", x = mean(c(-z, z)), y = 0.1, label = paste0(round(percentage_between, 2), "% between"), color = "blue") +
  coord_cartesian(xlim = c(-2.1, 2.1))+gonzo_theme()

```


```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(ggplot2)

# Set the Z-value
z <- as.numeric(mtcars_all_zscores[21,][12]) # Example Z-value

# Calculate the cumulative probabilities at the Z-values
cum_prob_positive <- pnorm(z)
cum_prob_negative <- pnorm(-z)

# Calculate the percentage between the Z-values (two-tailed)
percentage_between <- (cum_prob_positive - cum_prob_negative) * 100

# Generate a sequence of values along the x-axis
x <- seq(-3, 3, length.out = 1000)

# Calculate the corresponding probabilities under the standard normal curve
y <- dnorm(x)

# Create a data frame with x and y values for plotting
df <- data.frame(x, y)

# Create the plot with zoomed-in x-axis
ggplotsix<- ggplot(df, aes(x, y)) +
  geom_line() +
  geom_ribbon(data = subset(df, x >= -z & x <= z), aes(ymin = 0, ymax = y), fill = "blue", alpha = 0.3) +
  theme_minimal() +
  labs(x = "Z-value", y = "Density", title = "Percentage of Distribution Between +/- Z") +
  annotate("text", x = mean(c(-z, z)), y = 0.1, label = paste0(round(percentage_between, 2), "% between"), color = "blue") +
  coord_cartesian(xlim = c(-2.1, 2.1))+gonzo_theme()

library(gridExtra)

gridExtra::grid.arrange(ggplotfive,ggplotsix,ncol=2)
 

```

***
